---
title: "Lab 5 Assignment: Group 21"
author:
  - "Bertram B Broe (student id: ______)"
  - "Tóki / tokiboggle (student id: ______)"
  - "______ (student id: ______)"
  - "______ (student id: ______)"
format:
  html:
    embed-resources: true
editor: visual
---

*Fill in the blank student ids before upload. Render and zip the html for DTU Learn.*

# Background

Viral peptides get shown on the cell surface by MHC class I (HLA-A/B/C). CD8 T cells can then recognise the peptide–HLA complex via the TCR, especially the CDR3β loop. We used the ImmuneCODE MIRA class I tables (Nolan et al. 2020): `peptide-detail-ci.csv` and `subject-metadata.csv`.

# Aim

1. What characterises the peptides binding to the HLAs?
2. What characterises the T-cell receptors binding to the peptide–HLA complexes?

# Methods (short)

We loaded the two tables with `tidyverse` / `here`, kept class I alleles only, split `TCR BioIdentity` into CDR3b + V/J, exploded the comma-separated peptides into one row per peptide, dropped invalid amino-acid letters, and added sequence lengths. Alleles were cleaned to field 1+2 (e.g. `A02:01`) and joined to the peptide table on `Experiment`. Full step-by-step wrangling is in the group exercise file / `R/` scripts; here we only keep what we need for the two questions. Logos use `ggseqlogo`.

```{r}
#| label: setup
#| message: false
#| warning: false

library("tidyverse")
library("here")
library("ggseqlogo")

# Hand-in repo keeps raw files under data/_raw/ (gitignored).
# Use .gz if present, else the plain csv.
pep_gz <- here("data", "_raw", "peptide-detail-ci.csv.gz")
pep_csv <- here("data", "_raw", "peptide-detail-ci.csv")
peptide_data <- read_csv(
  if (file.exists(pep_gz)) pep_gz else pep_csv,
  show_col_types = FALSE
)
meta_data <- read_csv(
  here("data", "_raw", "subject-metadata.csv"),
  na = "N/A",
  show_col_types = FALSE
)

meta_data <- meta_data |>
  select(-starts_with("D")) |>
  rename(
    A1 = `HLA-A...9`,
    A2 = `HLA-A...10`,
    B1 = `HLA-B...11`,
    B2 = `HLA-B...12`,
    C1 = `HLA-C...13`,
    C2 = `HLA-C...14`
  ) |>
  select(Experiment, A1, A2, B1, B2, C1, C2) |>
  pivot_longer(cols = c(A1, A2, B1, B2, C1, C2),
               names_to = "Gene",
               values_to = "Allele") |>
  mutate(Allele = str_sub(Allele, 1, 7)) |>
  mutate(Allele = str_remove(Allele, "\\*")) |>
  filter(Allele != "", !is.na(Allele)) |>
  select(Experiment, Allele) |>
  distinct()

peptide_names <- str_c("peptide_", seq(1, 13))

peptide_data <- peptide_data |>
  select(`TCR BioIdentity`, Experiment, `Amino Acids`) |>
  separate(col = `TCR BioIdentity`,
           into = c("CDR3b", "V_gene", "J_gene"),
           sep = "\\+") |>
  separate(`Amino Acids`,
           into = peptide_names,
           sep = ",",
           fill = "right") |>
  pivot_longer(cols = starts_with("peptide_"),
               names_to = "peptide_n",
               values_to = "peptide") |>
  select(-peptide_n) |>
  drop_na(peptide) |>
  distinct() |>
  filter(
    str_detect(CDR3b, "[^ARNDCQEGHILKMFPSTWYV]", negate = TRUE),
    str_detect(peptide, "[^ARNDCQEGHILKMFPSTWYV]", negate = TRUE)
  ) |>
  mutate(
    k_CDR3b = str_length(CDR3b),
    k_peptide = str_length(peptide)
  )

peptide_meta_data <- peptide_data |>
  left_join(meta_data, by = "Experiment",
            relationship = "many-to-many") |>
  distinct()
```

# Analysis

## What characterises the peptides binding to the HLAs?

First we just looked at how long the peptides are:

```{r}
#| label: peptide-length
#| fig-cap: "Peptide lengths in the tidy table. Mode is 9."

peptide_data |>
  ggplot(aes(x = k_peptide)) +
  geom_bar(color = "black", alpha = 0.5) +
  theme_minimal() +
  labs(x = "Peptide length", y = "Counts")
```

Most peptides are length 9 (then 10). That matches what you usually hear for MHC-I, so the logos below are all 9-mers.

```{r}
#| label: logo-a0201
#| fig-cap: "Sequence logo of unique 9-mer peptides linked to allele A02:01."
#| warning: false
#| message: false

peptide_meta_data |>
  filter(Allele == "A02:01", k_peptide == 9) |>
  distinct(peptide) |>
  pull(peptide) |>
  ggseqlogo()
```

```{r}
#| label: logo-b0702
#| fig-cap: "Same idea for B07:02 (unique 9-mers)."
#| warning: false
#| message: false

peptide_meta_data |>
  filter(Allele == "B07:02", k_peptide == 9) |>
  distinct(peptide) |>
  pull(peptide) |>
  ggseqlogo()
```

**Answer.** The logos make it kinda clear that **positions 2 and 9** matter most for binding. Both have taller stacks than the middle. Position 2 is mostly **Leu (L)**, and position 9 is a hydrophobic pile with **Leu and Val** on top (plus some Y/F etc.). Positions 1 and 3–8 look pretty flat / mixed, so they don’t seem to drive the HLA binding as much.

The A02:01 and B07:02 logos look almost the same. That is a bit weird if you know classical B\*07:02 likes Pro at P2 — here we don’t really see that. One reason (also mentioned in the lab): the data only tells us the subject’s **haplotype**, not which allele actually presented the peptide. So when we filter on `B07:02` we also pull in peptides that might have been presented by another allele in the same person. Length and the “P2 + P9 anchors” story still show up; a clean allele-specific motif maybe doesn’t.

## What characterises the T-cell receptors binding to the peptide–HLA complexes?

CDR3b lengths first:

```{r}
#| label: cdr3-length
#| fig-cap: "CDR3b lengths. Mode is 15."

peptide_data |>
  ggplot(aes(x = k_CDR3b)) +
  geom_bar(color = "black", alpha = 0.5) +
  theme_minimal() +
  labs(x = "CDR3b length", y = "Counts")
```

Most common CDR3b length is **15**, so we picked that for the logo, together with allele `A02:01` and one peptide (`LLFLVLIML`), like in the exercises.

```{r}
#| label: logo-cdr3
#| fig-cap: "CDR3b logo for length 15, A02:01, peptide LLFLVLIML."
#| warning: false
#| message: false

peptide_meta_data |>
  filter(k_CDR3b == 15,
         Allele == "A02:01",
         peptide == "LLFLVLIML") |>
  distinct(CDR3b) |>
  pull(CDR3b) |>
  ggseqlogo()
```

**Answer.** The **ends** of the CDR3b are pretty fixed, and the **middle** is not. You basically always get **C** at the start and something like the **CASS…** stretch from the V gene, and the C-terminus piles up on **F** (from the J gene). In between (roughly positions 5–10) the letters are all over the place — that’s the junction, which is where a lot of the variability comes from when different TCRs contact the same peptide–HLA. So: ends set by V/J, middle junction variable.

Same haplotype caveat as above: we know the CDR3b was linked to that peptide in a subject who carries A02:01, not that A02:01 was proven to be the presenting allele.
