---
title: "Lab 5 Assignment: Group 21"
author:
  - Bertram B Broe
  - Tóki / tokiboggle
  - Name (student id)
  - Name (student id)
format:
  html:
    embed-resources: true
editor: visual
---

*Before upload: add the remaining names and student ids, render this file, and zip the html. Hand in on one course only.*

# Background

A viral peptide is a short scrap of a virus protein — basically a tiny leftover piece after the cell breaks the virus down. Those scraps get shown on the cell surface by MHC class I (Major Histocompatibility Complex class I), the display holders cells use for this job. In people those holders are also called HLA (Human Leukocyte Antigen). A CD8 T cell — a killer white blood cell — can recognize that peptide–HLA pair with its TCR (T-cell receptor), the sensor on the T cell that reads the name tag, especially CDR3 (complementarity-determining region 3), the sticky middle bit of the TCR that actually touches the pair. If the scrap looks viral, the T cell can tell the infected cell to die. That’s why we care.

We’re using the ImmuneCODE MIRA class I tables (Nolan et al. 2020), which link SARS-CoV-2 peptides to those TCRs and to each person’s HLA set, to ask what the peptides look like and what the TCRs look like when they show up together.

# Aim

Two questions, that’s it:

1. What characterises the peptides binding to the HLAs?
2. What characterises the T-cell receptors binding to the peptide–HLA complexes?

# Load libraries

```{r}
#| message: false
library(tidyverse)
library(ggseqlogo)
```

# Load data

The class I MIRA tables from Nolan et al. 2020. `peptide-detail-ci.csv` is the TCR–peptide hits. `subject-metadata.csv` is each person plus their HLA set. We join them on `Experiment`.

```{r}
#| message: false
peptide_raw <- read_csv(
  "data/_raw/peptide-detail-ci.csv",
  na = c("", "n/a", "N/A", "NA")
)

meta_raw <- read_csv(
  "data/_raw/subject-metadata.csv",
  na = c("", "n/a", "N/A", "NA")
)
```

# Data description

`peptide_raw` has one row per TCR hit, but the peptide column can hold several scraps in one cell, comma-separated. The TCR name is also a mash-up: CDR3b, then the V gene, then the J gene.

`meta_raw` has the people. We only keep class I holders (HLA-A, HLA-B, HLA-C). Each person has two copies of each, so we fold those into one `Allele` column and trim the names down to the usual A02:01-style label.

Then we stick the tables together on `Experiment`. One honest limit up front: this only says “this person had this HLA in their set.” It does not prove which holder actually showed the peptide. That matters later.

```{r}
#| message: false
aa_ok <- "ARNDCQEGHILKMFPSTWYV"

peptide_data <- peptide_raw |>
  select(`TCR BioIdentity`, Experiment, `Amino Acids`) |>
  separate(
    `TCR BioIdentity`,
    into = c("CDR3b", "V_gene", "J_gene"),
    sep = "\\+",
    extra = "merge",
    fill = "right"
  ) |>
  separate_rows(`Amino Acids`, sep = ",") |>
  rename(peptide = `Amino Acids`) |>
  filter(
    !is.na(CDR3b),
    !is.na(peptide),
    !str_detect(CDR3b, "unproductive"),
    !str_detect(CDR3b, str_c("[^", aa_ok, "]")),
    !str_detect(peptide, str_c("[^", aa_ok, "]"))
  ) |>
  mutate(
    k_CDR3b = str_length(CDR3b),
    k_peptide = str_length(peptide)
  ) |>
  distinct()

meta_data <- meta_raw |>
  select(Experiment, matches("^HLA-[ABC]")) |>
  rename(
    A1 = `HLA-A...9`,
    A2 = `HLA-A...10`,
    B1 = `HLA-B...11`,
    B2 = `HLA-B...12`,
    C1 = `HLA-C...13`,
    C2 = `HLA-C...14`
  ) |>
  pivot_longer(A1:C2, names_to = "copy", values_to = "Allele") |>
  filter(!is.na(Allele), Allele != "") |>
  mutate(
    Allele = Allele |>
      str_replace_all("\\*", "") |>
      str_replace("^([ABC])([0-9]{2}:[0-9]{2}).*$", "\\1\\2")
  ) |>
  select(Experiment, Allele) |>
  distinct()

peptide_meta_data <- peptide_data |>
  inner_join(meta_data, by = "Experiment", relationship = "many-to-many") |>
  distinct()
```

# Analysis

## Peptides

We took unique 9-mers from people who carry A02:01, and the same for B07:02. Nine is the usual scrap length for these holders. The plot is a sequence logo: tall letters are the amino acids that keep showing up at that spot.

```{r}
#| fig-width: 8
#| fig-height: 3
#| fig-cap: "Unique 9-mer peptides from people who carry HLA-A02:01."
#| warning: false
a02_9mers <- peptide_meta_data |>
  filter(Allele == "A02:01", k_peptide == 9) |>
  distinct(peptide) |>
  pull(peptide)

ggseqlogo(a02_9mers, seq_type = "aa")
```

```{r}
#| fig-width: 8
#| fig-height: 3
#| fig-cap: "Unique 9-mer peptides from people who carry HLA-B07:02."
#| warning: false
b07_9mers <- peptide_meta_data |>
  filter(Allele == "B07:02", k_peptide == 9) |>
  distinct(peptide) |>
  pull(peptide)

ggseqlogo(b07_9mers, seq_type = "aa")
```

What jumps out is not the middle. Positions 2 and 9 do the work. Position 2 likes leucine. Position 9 likes oily leftovers — leucine, valine, tyrosine. Those two spots are the feet that sit in the holder. The middle of the scrap can wander.

The two logos also look almost the same. That is the haplotype caveat again: we tagged every peptide with every class I allele the person had. So this is “scraps from people who own A02:01,” not “scraps we watched A02:01 hold.” Still, the two-point grip is the thing that characterises the peptides.

## CDR3 sequences

Same idea, now for the TCR side. We kept CDR3b sequences of length 15 that show up with A02:01 and the peptide LLFLVLIML.

```{r}
#| fig-width: 8
#| fig-height: 3
#| fig-cap: "Length-15 CDR3b sequences seen with A02:01 and peptide LLFLVLIML."
#| warning: false
cdr3_15 <- peptide_meta_data |>
  filter(
    Allele == "A02:01",
    k_CDR3b == 15,
    peptide == "LLFLVLIML"
  ) |>
  distinct(CDR3b) |>
  pull(CDR3b)

ggseqlogo(cdr3_15, seq_type = "aa")
```

The ends are boring on purpose. They come with the V and J gene pieces, so you get CASS at the start and a phenylalanine at the end, often with EQY or similar just before it. The middle is the junction — that is the bit that actually leans on the peptide–HLA pair, and it is much more mixed.

Same caveat as above: we know the person had A02:01. We do not know that A02:01 was the holder for LLFLVLIML.

So: peptides are characterised by a two-point grip (leucine at 2, something oily at 9). TCRs are characterised by fixed V/J ends and a wiggly middle that reads the pair.
